\vspace{1.0cm} \huge \bf Supplemental information:\\ \LARGE Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-08-07 fluB.tex
Fix #40, #39
Clarify deltaTMRCA confusion. I've decided that deltaTMRCA is best
thought of as the mean of absolute deviations in TMRCAs between 2
trees. We get normalized deltaTMRCA by quantifying deltaTMRCAs within
(=phylogenetic noise) and between (=phylogenetic noise + signal) trees.
The result of this is a measure of how similar the signal is to noise,
which in our case is the desired outcome. Hopefully that makes sense.
Also move all the LD figures to supplementary. It was done as a
confirmation of results and also addresses the figure number problem.

\subsection*{Analysis of reassortment properties}
We attempted to quantify the temporal discordance between lineages reassorting into new genomic constellations.
If one were able to recover an influenza `species tree', including admixture/reassortment events, it would be possible to estimate the reassortment or recombination `distance', which is the time between a split in the species tree in the past and a reassortment event (see Figure \ref{speciesTree}).
Although we do not find evidence of differences in total number of reassortments between segments (see Supplementary information), we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
This difference in TMRCAs, or $\dtmrca$, is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
We normalize our $\dtmrca$ comparisons to account for uncertainty in tree topology (see Methods).
Figure \ref{deltaTMRCA} shows normalized mean $\dtmrca$ values for all pairs of trees.
Most segment pairs show very low values for this statistic with $\dtmrca \approx 0.1$, indicating that $\dtmrca$ measurements between replicate posterior samples from the same segment are up to 10 times smaller than $\dtmrca$ values between posterior samples from different segments.
PB1, PB2 and HA trees, on the other hand, exhibit normalized $\dtmrca$ values that are much higher.
This shows that TMRCA differences between trees of PB1, PB2 and HA segments are, though noisy, occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.

@@

We next sought to confirm our findings through measurement of linkage disequilibrium (LD), a measure of non-random association between polymorphic loci within a population.
We estimated LD directly from haplotype frequencies at polymorphic amino acid sites (see Methods), thereby avoiding phylogenetic reconstruction or Vic/Yam lineage assignment.
We observe greater amino acid LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}) in a large secondary dataset.
This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting non-synonymous alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
We conclude that Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions.
Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 4 sites on PB1, 4 on PB2 and 4 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
In addition, there are sites on PB1, PB2, HA and NA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.

\begin{figure}[h]
	\centering	
	\includegraphics[width=0.65\textwidth]{figures/InfB_LDcircle.1600.aa.ChiSqdf.minorCutoff1percent.png}
	\caption{\textbf{LD comparison between influenza B proteins.}
Pairwise comparisons of linkage disequilibrium between amino acid sites on influenza B proteins in the secondary dataset.
Many polymorphic amino acid sites on PB1, PB2 and HA proteins exhibit high LD between themselves, followed by the NA protein.
This is evidence of a considerable number of co-assorting alleles within these proteins.}
	\label{segmentLD}
\end{figure}

@@

\subsection*{Tree to tree similarities}
We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ for a particular posterior sample $i$, following
\begin{equation}
\dtmrca(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
\end{equation}
where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $n$ is the total number of pairwise comparisons available between sets of tips.
Thus, $g(A_{ij},B_{ij})$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree $A$ and the \textit{i}th posterior sample of tree $B$.
Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), to control for variability in tree topology stability over the course of the MCMC chain.
We had 3 replicate analyses of each segment and in order to calculate $f(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
We subsampled our combined posterior distribution of trees to give a total of 2,700 trees on which to analyze $\dtmrca$.

Calculating $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
Our approach exploits the branch scaling used by BEAST \citep{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
In the absence of reassortment we expect the tree of every segment to recapitulate the `virus tree', a concept analogous to `species trees' in population genetics.
Our method operates under the assumption that the segment trees capture this `virus tree' of influenza B viruses quite well.
It is not an unreasonable assumption, given the seasonal bottlenecks influenza viruses experience.
This makes it almost certain that influenza viruses circulating at any given time point are derived from a single genome that existed in the recent past.
The $\dtmrca$ statistic essentially quantifies the temporal distance between admixture events and nodes in the `virus tree' (see Figure \ref{speciesTree}).
Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \citep{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \citep{ruane2011}.

>>>>

\subsection*{Analysis of reassortment properties}
We attempted to quantify the temporal discordance between lineages reassorting into new genomic constellations.
If one were able to recover an influenza `species tree', including admixture/reassortment events, it would be possible to estimate the reassortment or recombination `distance', which is the time between a split in the species tree in the past and a reassortment event (see Figure \ref{speciesTree}).
Although we do not find evidence of differences in total number of reassortments between segments (see Supplementary information), we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
The summary statistic we use that reflects this difference in TMRCAs, $\dtmrca$, is most sensitive when only one of the two trees being compared loses diversity via reassortment and the other acts like a proxy for the `species tree'.
We normalize our $\dtmrca$ comparisons to account for uncertainty in tree topology (see Methods).
Figure \ref{deltaTMRCA} shows normalized $\dtmrca$ values for all pairs of trees.
Most segment pairs show very low values for this statistic with $\dtmrca \approx 0.1$, indicating that $\dtmrca$ measurements between replicate posterior samples from the same segment are up to 10 times smaller than $\dtmrca$ values between posterior samples from different segments.
PB1, PB2 and HA trees, on the other hand, exhibit normalized $\dtmrca$ values that are much higher.
This shows that TMRCA differences between trees of PB1, PB2 and HA segments are, though noisy, occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.

@@

@@

\subsection*{Tree to tree similarities}
We express the distance $\dtmrca$ between trees belonging to two segments $A$ and $B$ for a particular posterior sample $i$, following
\begin{equation}
normalized \dtmrca(A_i, B_i) = \frac{\dtmrca(A_i, A_i') + \dtmrca(B_i, B_i')}{2 \, \dtmrca(A_i, B_i)},
\end{equation}
where $\dtmrca(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $n$ is the total number of pairwise comparisons available between sets of tips.
Thus, $g(A_{ij},B_{ij})$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree $A$ and the \textit{i}th posterior sample of tree $B$.
Additionally, $\dtmrca(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree $A$ (which we refer to as $A'$), which is used in the normalization procedure to control for variability in tree topology stability over the course of the MCMC chain (see Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates}).
We had 3 replicate analyses of each segment and in order to calculate $\dtmrca(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as $A$ and analyses numbered 2, 3 and 1 as $A'$, in that order.
We subsampled our combined posterior distribution of trees to give a total of 2,700 trees on which to analyze normalized $\dtmrca$.

Calculating normalized $\dtmrca(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this statistic allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
Our approach exploits the branch scaling used by BEAST \citep{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.
In the absence of reassortment we expect the tree of every segment to recapitulate the `virus tree', a concept analogous to `species trees' in population genetics.
Our method operates under the assumption that the segment trees capture this `virus tree' of influenza B viruses quite well.
It is not an unreasonable assumption, given the seasonal bottlenecks influenza viruses experience.
This makes it almost certain that influenza viruses circulating at any given time point are derived from a single genome that existed in the recent past.
The $\dtmrca$ statistic essentially quantifies the temporal distance between admixture events and nodes in the `virus tree' (see Figure \ref{speciesTree}).
Our $\dtmrca$ statistic is an extension of patristic distance methods and has previously been used to tackle a wide variety of problems, as phylogenetic distance in predicting viral titer in \textit{Drosophila} infected with viruses from closely related species \citep{longdon2011} and to assess temporal incongruence in a phylogenetic tree of amphibian species induced by using different calibrations \citep{ruane2011}.

### 2014-08-07 fluB_supp.tex
Fix #40, #39
Clarify deltaTMRCA confusion. I've decided that deltaTMRCA is best
thought of as the mean of absolute deviations in TMRCAs between 2
trees. We get normalized deltaTMRCA by quantifying deltaTMRCAs within
(=phylogenetic noise) and between (=phylogenetic noise + signal) trees.
The result of this is a measure of how similar the signal is to noise,
which in our case is the desired outcome. Hopefully that makes sense.
Also move all the LD figures to supplementary. It was done as a
confirmation of results and also addresses the figure number problem.

>>>>

\subsection*{Confirmation of primary findings}
We sought to confirm our findings through measurement of linkage disequilibrium (LD), a measure of non-random association between polymorphic loci within a population.
We estimated LD directly from haplotype frequencies at polymorphic amino acid sites (see Methods) in the secondary dataset, thereby avoiding phylogenetic reconstruction or Vic/Yam lineage assignment.
We observe greater amino acid LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}) in a large secondary dataset.
This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting non-synonymous alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
We conclude that Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions.
Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 4 sites on PB1, 4 on PB2 and 4 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
In addition, there are sites on PB1, PB2, HA and NA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.

\begin{figure}[h]
	\centering	
	\includegraphics[width=0.85\textwidth]{supp_figures/InfB_LDcircle.1600.aa.ChiSqdf.minorCutoff1percent.png}
	\caption{\textbf{LD comparison between influenza B proteins.}
Pairwise comparisons of linkage disequilibrium between amino acid sites on influenza B proteins in the secondary dataset.
Many polymorphic amino acid sites on PB1, PB2 and HA proteins exhibit high LD between themselves, followed by the NA protein.
This is evidence of a considerable number of co-assorting alleles within these proteins.}
	\label{segmentLD}
\end{figure}
